The Effect of Oil Price on Field Production: Evidence from the Norwegian Continental Shelf

### 2016-03-04 oil_prices_obes.tex
update draft

\documentclass[12pt]{article}
\usepackage[margin=1in]{geometry}
\usepackage{setspace}
\usepackage{graphicx}
\usepackage{amsmath}
\usepackage{natbib} %for citet and citep
\usepackage{syntonly}
\usepackage{esdiff} %for writing partial derivatives
\usepackage{url} %for inserting urls
\usepackage{placeins} % for limitting floats
\usepackage{comment} %for excluding figures
% \excludecomment{figure} % for printing without figures
% \syntaxonly % for quickly checking document
%set document settings

@@

Oil and gas supply, especially in difficult and expensive offshore areas, is the production of an often lengthy and complex chain of conditional investment decisions. Starting with the exploration phase, where a potential oil-producing region or "play" is explored, leases are acquired, prospects are identified and exploration wells are drilled. In the case of a significant hydrocarbon find, a field is appraised to judge size, difficulty and in turn profitability. For fields seen as sufficiently profitable, a development stage commences, where a detailed drilling and production plan is drawn up and submitted to obtain permits and regulatory approval. Operating agreements are signed and subcontractors acquired for offshore construction and well development. Finally, at the end of this multi-year process, a build-out can commence, where full production capacity is built out as fast as possible, followed by an extended depletion phase. All of this consists of what is referred to as upstream activity, with steps such as shipping, refining and trading, among other, making up an equally complex down-stream process. \footnote{\ref{leffler_deepwater} provides an accessible guide to complex offshore petroleum exploration and production industry.}

@@

\begin{figure}
	\includegraphics[width=1\textwidth]{figures/investment.png}
	\caption{The figure shows investments (millions NOK) in the oil and gas extraction industry on the Norwegian Continental Shelf. A large and growing share of investment goes into fields that are already producing as opposed to investments in search, exploration and build-out of new fields.}
	\label{investment}
\end{figure}

@@

A similar analysis combining econometric models with shape-fitting models by \citet{cleveland_forecasting_1991} also led to pessimistic forecasts of future US production. Further refinements of these were done by, among others, \citet{pesaran_forecasting_1995}  and \citet{moroney_integrated_1999}. \citet{kaufman_oil_2001} present an extension and generalisation using a Vector Error Correction Model (VECM) that relaxes the assumption that firms produce their fields in order of decreasing quality and also includes institutional factors in US oil production. They claim that they are able to account for most of the price variation between 1938 and 1991.

\citet{adelman_mineral_1990} and \citet{adelman_modeling_1993} take an opposing view, using cointegration and causality testing to argue that scarcity, and in turn geophysical forces had little to do with the eras oil price movements. Instead, Adelman argued that cartel behavior of low cost producers likely stood behind the increasing oil prices. \citet{lynch_forecasting_2002} provides a fuller overview of these competing strands of the litterature.

@@

In addition to data on field-level production, I also have access to data on estimates of total recoverable reserves as well as estimates of total in-place oil for each field. The use of total recoverable reserves is complicated since it is dependent on the price of oil and thus endogenous in the analysis.  Therefore, as a proxy for the size of the fields, I use estimates of in-place oil from the Norwegian Petroleum Directorate, which is unlikely to be correlated with prices. The left panels of figure \ref{data_descriptives} show the histograms of field sizes as measured by estimates of recoverable oil and in-place oil.

@@

@@

\begin{equation}
\begin{split}
	Log(production_{i,t}) & = f(production\_time_{i,t}) + f(year_t) + \beta_1 in\_place\_oil_{i,t}\\
	 \quad & + \beta_3 cost\_index_{t} + \mathbf{\beta_2 oil\_price\_lags_t} +  
	  \epsilon_{i,t}
\label{gam_price_eqn}
\end{split}
\end{equation}

In this model I am estimating the parameters and functions from all fields $i$. The left-hand-side variable, $Log(production_{i,t})$ is the log of yearly oil production for field $i$ at production time $t$ in millions of standard cubic meters (SM3).  On the right hand side is a smoothed function of the production profile over time $f(production\_time_{i,t})$; $in\_place\_oil_{i,t}$ is the estimated in-place oil of the field, and $cost\_index_{t}$ represents the producer price index for oil and gas extraction.

>>>>

\documentclass[12pt]{article}
\usepackage[margin=1in]{geometry}
\usepackage{setspace}
\usepackage{graphicx}
\usepackage{amsmath}
\usepackage{natbib} %for citet and citep
\usepackage{syntonly}
\usepackage{esdiff} %for writing partial derivatives
\usepackage{url} %for inserting urls
\usepackage{placeins} % for limitting floats
\usepackage{comment} %for excluding figures
\usepackage{booktabs}
% \excludecomment{figure} % for printing without figures
% \syntaxonly % for quickly checking document
%set document settings

@@

Oil and gas supply, especially in difficult and expensive offshore areas, is the production of an often lengthy and complex chain of conditional investment decisions. Starting with the exploration phase, where a potential oil-producing region or "play" is explored, leases are acquired, prospects are identified and exploration wells are drilled. In the case of a significant hydrocarbon find, a field is appraised to judge size, difficulty and in turn profitability. For fields seen as sufficiently profitable, a development stage commences, where a detailed drilling and production plan is drawn up and submitted to obtain permits and regulatory approval. Operating agreements are signed and subcontractors acquired for offshore construction and well development. Finally, at the end of this multi-year process, a build-out can commence, where full production capacity is built out as fast as possible, followed by an extended depletion phase. All of this consists of what is referred to as upstream activity, with steps such as shipping, refining and trading, among others, making up an equally complex down-stream process. \footnote{\citet{leffler_deepwater_2011} provides an accessible guide to the offshore petroleum exploration and production industry.}

@@

\begin{figure}
	\center
	\includegraphics[width=.9\textwidth]{figures/investment.png}
	\caption{The figure shows investments (millions NOK) in the oil and gas extraction industry on the Norwegian Continental Shelf. A large and growing share of investment goes into fields that are already producing as opposed to investments in search, exploration and build-out of new fields.}
	\label{investment}
\end{figure}

@@

A similar analysis combining econometric models with shape-fitting models by \citet{cleveland_forecasting_1991} also led to pessimistic forecasts of future US production. Further refinements of these were done by, among others, \citet{pesaran_forecasting_1995}  and \citet{moroney_integrated_1999}. \citet{kaufmann_oil_2001} present an extension and generalisation using a Vector Error Correction Model (VECM) that relaxes the assumption that firms produce their fields in order of decreasing quality and also includes institutional factors in US oil production. They claim that they are able to account for most of the price variation between 1938 and 1991.

\citet{adelman_mineral_1990} and \citet{adelman_modelling_1993} take an opposing view, using cointegration and causality testing to argue that scarcity, and in turn geophysical forces had little to do with the eras oil price movements. Instead, Adelman argued that cartel behavior of low cost producers likely stood behind the increasing oil prices. \citet{lynch_forecasting_2002} provides a fuller overview of these competing strands of the litterature.

@@

In addition to data on field-level production, I also have access to data on estimates of total recoverable reserves as well as estimates of total in-place oil for each field. The use of total recoverable reserves is complicated since it is dependent on the price of oil and thus endogenous in the analysis. Therefore, as a proxy for the size of the fields, I use estimates of geological in-place oil which is not a function of economic factors. The left panels of figure \ref{data_descriptives} show the histograms of field sizes as measured by estimates of recoverable oil and in-place oil.

@@

Table \ref{field_summary} provides summary statistics over the distribution of field-level variables. Notably, the median field in the data set had been producing for 16 years as of the beginning of 2015 - an indication of the mature status of the Norwegian Continental Shelf.

\begin{table}
\begin{tabular}{lrrrrr}
\toprule
Percentiles &    5\% &    25\% &    50\% &     75\% &     95\% \\
                      &       &        &        &         &         \\
\midrule
In-place Oil, Mill SM3         &  5.90 &  34.00 &  75.80 &  157.00 &  636.70 \\
Est. Recoverable Oil           &  6.20 &  10.50 &  25.70 &   70.30 &  374.50 \\
Production Years, to 2015      &  1.79 &   9.17 &  16.04 &   22.43 &   35.63 \\
Tot. Cumulative Prod., to 2015, Mill SM3 &  3.61 &   8.94 &  22.94 &   57.46 &  359.04 \\
\bottomrule
\end{tabular}
\caption{Summary distribution statistics of field-level data. Columns represent percentiles of data. Excluds the Ekofisk field and fields with less than 5 million SM3 of recoverable oil.}
\label{field_summary}
\end{table}

@@

\begin{equation}
\begin{split}
	Log(production_{i,t}) & = f(production\_time_{i,t}) + f(year_t) + \beta_1 in\_place\_oil_{i}\\
	 \quad & + \beta_3 cost\_index_{t} + \mathbf{\beta_2 oil\_price\_lags_t} +  
	  \epsilon_{i,t}
\label{gam_price_eqn}
\end{split}
\end{equation}

In this model I am estimating the parameters and functions from all fields $i$. The left-hand-side variable, $Log(production_{i,t})$ is the log of yearly oil production for field $i$ at production time $t$ in millions of standard cubic meters (SM3).  On the right hand side is a smoothed function of the production profile over time $f(production\_time_{i,t})$; $in\_place\_oil_{i}$ is the estimated in-place oil of the field, and $cost\_index_{t}$ represents the producer price index for oil and gas extraction.
